---
title: "NBA Draft Network Analysis Project"
author: "Marcus Li and Andrew Ung"
date: "6/6/2025"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    theme: readable
    highlight: tango
---

## Narrative Summary

This project analyzes NBA player career trajectories and teammate networks using network analysis. We focus on the \_\_\_ draft class and \_\_\_ season teammates to explore how players connect through shared teams, and use network metrics and community detection to identify key players and roles. We also will be exploring "broker" players The goal is to visualize and interpret player dynamics through graph structures.

We use `nbastatR` to access real NBA data, and `igraph`, `ggraph`, and `dplyr` for data wrangling, network creation, and visualization. This work could help illustrate team structures and player centrality to coaches, analysts, or fans. \## Setup Process

```{r setup}
#installing packages
install.packages("remotes")
remotes::install_github("abresler/nbastatR", force = TRUE)

```

### Packages Used

```{r packages}
#packages needed
library(devtools)
library(dplyr)
library(future)
library(ggplot2)
library(igraph)
library(nbastatR)
library(Rtools)

library(sna)  # For brokerage analysis
```

### Reading the Data Sets

```{r datasets}
#Adjust VROOM buffer size for large data!
Sys.setenv(VROOM_CONNECTION_SIZE = 131072 * 10)

#change season to anything from (ex 2020:2024, from 2020 to 2024 seasons)
game_data <- game_logs(seasons = 2000:2010)
```

### Creating Graph Objects

```{r graphObjects}
# Create player-to-player network based on shared teams
cat("Creating player teammate network...\n")


# Get all player-team-season combinations
player_teams <- game_data %>%
  select(namePlayer, slugTeam, slugSeason) %>%
  distinct()
  
teammate_pairs <- player_teams %>%
  group_by(slugTeam, slugSeason) %>%
  filter(n() > 1) %>%
  do({
    players <- .$namePlayer
    expand.grid(
      player1 = players,
      player2 = players,
      stringsAsFactors = FALSE
    ) %>%
    filter(player1 != player2)
  }) %>%
  group_by(player1, player2) %>%
  summarise(
    shared_teams = n(),
    teams = paste(unique(paste(slugTeam, slugSeason)), collapse = ", "),
    .groups = "drop"
  ) %>%
  filter(shared_teams >= 1)

cat("Teammate connections created:", nrow(teammate_pairs), "\n")

```

### Filtering

```{r filter}
#connection between players
player_connections <- teammate_pairs %>%
  group_by(player1) %>%
  summarise(total_teammates = n(), .groups = "drop") %>%
  arrange(desc(total_teammates))

#choose the top 50%
top_players <- head(player_connections$player1, 100)

filtered_network <- teammate_pairs %>%
  filter(player1 %in% top_players & player2 %in% top_players)


#Creates graph
players_network <- graph_from_data_frame(filtered_network, directed = FALSE)
# Simplify graph (remove loops and multiple edges)
players_network <- simplify(players_network)
```


###Detection

```{r detetcion}
# Community detection
# Example in igraph
communities <- cluster_louvain(players_network)


V(players_network)$community <- membership(communities_louvain)

# Betweenness centrality
between <- igraph::betweenness(players_network)


#showcases top broker players with between centrality
at("=== (Betweenness Centrality) ===\n")
print(head(bet.dat[order(bet.dat$between, decreasing = TRUE),], 10))

#community structure
cat("\n=== COMMUNITY STRUCTURE ===\n")
cat("Number of communities:", max(membership_vec), "\n")
cat("Modularity:", round(modularity(communities), 3), "\n")

#members of each community
for(i in 1:min(5, max(membership_vec))) {
  community_players <- V(players_network)$name[membership_vec == i]
  cat("Community", i, ":", paste(head(community_players, 5), collapse = ", "), 
      ifelse(length(community_players) > 5, "...", ""), "\n")
}
```

### Brokerage Analysis
```{r brokerage}
library(intergraph)
library(statnet)
library(knitr)
library(dplyr)

# Convert igraph to network
snaNetwork <- asNetwork(players_network)

# Confirm the network size matches igraph vertex count
stopifnot(network.size(snaNetwork) == length(V(players_network)))

community_named <- membership(communities)
names(community_named) <- V(players_network)$name

# 5. Assign community attribute to the network vertices by matching names
set.vertex.attribute(
  snaNetwork,
  "community",
  value = community_named[get.vertex.attribute(snaNetwork, "vertex.names")]
)

# 6. Confirm community distribution
print(table(get.vertex.attribute(snaNetwork, "community")))

# 7. Run brokerage analysis
brokerage_results <- brokerage(snaNetwork, cl = get.vertex.attribute(snaNetwork, "community"))

# Create dataframe from results
brokerage_df <- round(brokerage(snaNetwork, cl=get.vertex.attribute(snaNetwork, "community"))$z.nli, 2) 

#un rounded non nom,rialized
#brokerage_dftest <- brokerage_results$raw.nli
colnames(brokerage_df) <- c(
  "Coordinator",     # was w_I
  "Consultant",      # was w_O
  "Representative",  # was b_IO
  "Gatekeeper",      # was b_OI
  "Liaison",         # was b_O
  "Total"            # was t
)

# Show all data sorted by Total
print("=== ALL PLAYERS BY TOTAL BROKERAGE ===")
print(brokerage_df[order(brokerage_df[, "Total"], decreasing = TRUE), ])

# Show top 5 by Total
print("=== TOP 5 TOTAL BROKERS ===")
print(head(brokerage_df[order(brokerage_df[, "Total"], decreasing = TRUE), ], 5))

# Show significant brokers (z > 1.96)
significant <- brokerage_df[brokerage_df[, "Total"] > 1.96, ]
if(nrow(significant) > 0) {
  print("=== SIGNIFICANT TOTAL BROKERS (z > 1.96) ===")
  print(significant[order(significant[, "Total"], decreasing = TRUE), ])
}

# Show top player in each category
print("=== TOP PLAYER BY CATEGORY ===")
for(col in colnames(brokerage_df)) {
  top_player <- rownames(brokerage_df)[which.max(brokerage_df[, col])]
  top_score <- max(brokerage_df[, col])
  print(paste(col, ":", top_player, "(", top_score, ")"))
}
```

### Description of the Data and Link

## Network Plot

```{r plotting}
# Find the top player in each community
community_tops <- by(1:length(V(players_network)), membership(communities), function(idx) {
  community_betweenness <- between[idx]
  idx[which.max(community_betweenness)]
})

# Convert to vector of indices
top_in_each_community <- unlist(community_tops)

# Create label color and font vectors
label_colors <- rep("black", length(V(players_network)))
label_fonts <- rep(1, length(V(players_network)))

# Top player in each community - red and bold
label_colors[top_in_each_community] <- "blue"
label_fonts[top_in_each_community] <- 2

# Plot network
plot(players_network,
     layout = layout_with_kk(players_network),
     vertex.label = ifelse(between > quantile(between, 0.1), 
                           V(players_network)$name, NA),
     vertex.label.color = label_colors,
     vertex.label.font = label_fonts,
     vertex.label.cex = 0.8,
     vertex.color = membership(communities),
     vertex.size = between * 0.01 + 3,
     edge.width = 0.1,
     edge.color = "gray80",
     main = "NBA Player Communities\n(Red Labels = Community Leaders, Colors = Communities)")

```

## Final Summary
